# Logistic regression

Up to now you have practiced with hypothese test for numerical continuous data, but in some cases the data behaves differentially.
A common type of data is binomial: categorical data with only two outcomes.

For example, a tomato grower is experimenting with the best time to harvest his tomatos:

* On six plants all tomatoes are harvested 10 days before the normal harvest time
* On four plants all tomatoes are harvested 5 days before the normal harvest time
* On another six plants all tomatoes are harvested at the normal harvest time

All harvested tomatoes are labelled ripe or not ripe:

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(readxl)
library(tidyverse)
theme_set(theme_classic(base_size = 16))

tomaat <- read_excel('../data/tomaten.xlsx')
knitr::kable(table(tomaat$tijd, tomaat$rijp))


```

Zulke data kan je mooi presenteren in een mozaiekplot (In het Engels *mosaic plot* geheten):
This kind of data can be clearly presented in a mosaic plot (see script for inspiration):

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(ggmosaic)

tomaat %>% 
  ggplot() +
  geom_mosaic(aes(x=product(tijd), fill = rijp)) +
  scale_fill_manual(values = c("green", "red")) +
  xlab("tijd (days before harvest)") +
  ylab("fraction ripe") +
  theme(legend.position = "none")

```

A nice feature of a mosaic plot is that you can read on both x- and y-axis which fraction is found in each category.

::: {#exr-ggmosaic}

mosaic plots

* install the package ggmosaic
* Import the data set tomaten.xlsx
* Make a mosaic plot with above code
* try to find out what each line of code is doing, writh down comments

:::

## binomial alternative for ANOVA's and linear regressions

The normal **linear model** cannot handle a binomial respons veriable.
For that you need to use a **generalized linear model**.
The function in R is `glm()`.
You can add your statistical model in the same way as in the function `lm()`, but you have to add an extra argument to make clear that your respons is binomial: `family = binomial()`.

::: {.callout-warning}

The response variable needs to be either a numerical variable with two values, or a factor (use `factor()`)

:::

> H~0~: all harvest times have the same probability for ripe tomatoes

> H~1~: at least some harvest times differ in probability for ripe tomatoes


```{r, echo=TRUE, message=FALSE, warning=FALSE}
fit <- glm(factor(rijp)~factor(tijd), family = binomial(), data = tomaat)

```

NB: the function `factor()` is used for the explanatory variable to make clear to R that the numerica data needs to be treated as categorical data. 

The normal `anova()` function cannot handle the output of `glm()`.
The reason is that you cannot apply a normal analysis of variance on these type of models to calculate the p-values.
Instead you need to use the `Anova()` function from the car package:


```{r, echo=TRUE, message=FALSE, warning=FALSE}
car::Anova(fit)
```


::: {.callout-warning}

The package car contains a number of functions with the same name as functions in tidyverse.
Therefore it is not practical to activate all functions.
Instead use the code `car::Anova()` to make R clear in which package the function can be found.

:::

::: {#exr-carpackage}

car package

* Install the car packages

:::


## Posthoc test

Similar as in normal ANOVA's you can apply a posthoc test with the function `emmeans()`


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(emmeans)
emmeans(fit, specs = pairwise ~ tijd)

```


## Practice

A third year student Applied Biology is doing her internship at a grower and needs to test five different insecticides.
She designed the following experiment:

* A certain number of flies is exposed to one of the five insecticides
* after one hour she counts the number of dead and still alive flies

You find the data set in your data map: insecticiden.xlsx

::: {#exr-insectices}

Insecticides.


* import the data set
* Make a nice graph
* apply a logistic regression
* Test which insecticide kills significant more flies than the control

:::


## Regression with binomial data

We can also present the data in a different way:

```{r echo=TRUE, message=FALSE, warning=FALSE}
tomaat %>% 
  ggplot(aes(tijd, rijp, color = rijp)) + 
  geom_count() +
  scale_color_manual(values = c("green", "red")) 
```

With the function `geom_count()` completely overlapping points are represented with bigger dots.
Alternative is `geom_jitter()` that randomly move the points a bit to reduce overlap.

Now we have on the x-axis not categories, but a continuous scale.

Can we do something with regression?
The answer is yes!

As with `lm()` the function `glm()` can perform both ANOVAs and regressions.
This time the regression is not linear, but logistic: a **logistic regression**.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
fit2 <- glm(factor(rijp) ~ tijd, family = binomial(), data = tomaat)
```

The only difference with the previous analysis is that now the variable tijd is NOT a factor.


```{r, echo=TRUE, message=FALSE, warning=FALSE}
car::Anova(fit2)

```


For the tomato grower it is important to know from what time at least 25% of the tomatoes is ripe.

We can visualise that by adding a trend line (this time `method = glm`) and a horizontal line through 0.25:

```{r echo=TRUE, message=FALSE, warning=FALSE}
tomaat %>% 
  mutate(rijp = recode(rijp, rijp=1, onrijp=0)) %>%
  ggplot(aes(tijd, rijp)) + 
  geom_count() +
  geom_smooth(method=glm, method.args = list(family = binomial())) +
  ylab("Fractie rijp") +
  geom_hline(yintercept = 0.25, color = "blue", linetype = "dashed") +
  theme(legend.position = "none") 
```

Now you see the typical S-shape of a logistic regression.
From five days before the normal harvest data the fraction ripe tomatoes is above the threshold.

::: {.callout-warning}

Within ggplot the function `glm()` cannot handle a factor response, but only 0 and 1 values. With the `recode` you can change your text to other values.

:::

::: {#exr-guppies}

Guppies


Pitkow *et al.* (1960) studied the effect of exposure to low temperature on the survival of guppies.
He exposed each time 40 guppies to water at 5°C, for 3, 8, 12 or 18 minutes. 

* import the data set (guppies.xlsx)
* make a clear graph, add a trend line
* apply a logistic regression
* what are your conclusions

:::

